A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2021-12-22 getdata3.Rmd
further development based on skum::pers

```{r fromobs}
bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
indx <- match(per$id, bobs$id)
per$nofrs <- bobs$nofrs[indx]
per <- per[!is.na(per$nofrs), ]
per$subreg <- factor(per$nofrs)
levels(per$subreg) <- c(rep("ume", 2), 
                        rep("bjur", 4), 
                        rep("sten", 3))
per <- per[(per$subreg != "sten"), ]
per$subreg <- factor(per$subreg)
per$enter <- 0
saveRDS(per, file = "data/infdat3.rds")
```

>>>>

```{r fromobs}
bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
indx <- match(per$id, bobs$id)
per$nofrs <- bobs$nofrs[indx]
per <- per[!is.na(per$nofrs), ]
per$subreg <- factor(per$nofrs)
levels(per$subreg) <- c(rep("ume", 2), 
                        rep("bjur", 4), 
                        rep("sten", 3))
per <- per[(per$subreg != "sten"), ]
per$subreg <- factor(per$subreg)
per$enter <- 0
per$sex <- factor(per$kon, labels = c("boy", "girl"))

##saveRDS(per, file = "data/infdat3.rds")
```

# Put on temperature data

```{r makecommunal}
temp_start <- 1894
temp_end <- 1951
temp_ivl <- c(temp_start, temp_end)
##
umetemp <- readRDS("data/umetemp.rds")
bjurtemp <- readRDS("data/bjurtemp.rds")
##

lagg <- 3 / 365  ## NOTE!!
tempvars <-  c("heat", "aver", "cold", "emintemp", "emeantemp", "emaxtemp", 
               "week", "year") 
## Ume:
bume <- per[per$subreg == "ume", ]
bume <- make.communal(bume, umetemp[, tempvars], 
                      start = temp_start, period = 1/52, lag = lagg)

### Ske:
bskel <- per[per$subreg == "bjur", ]
bskel <- make.communal(bskel, bjurtemp[, tempvars], 
                       start = temp_start, period = 1/52, lag = lagg)

bume$region <- "ume"
bskel$region <- "ske"
infdat <- rbind(bume, bskel)
infdat <- infdat[order(infdat$id, infdat$enter), c("region", "id", "sex", "enter", "exit", "event", 
                                                   "birthdate", "socStatus", "hisclass", "illeg",
                                                   "parity", tempvars)]
saveRDS(infdat, file = "data/infdat2_year2.rds")

```

### 2021-12-22 paper.Rmd
modifications positively ...
further development based on skum::pers

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + 365 * x))^3
ginv <- function(y) (1 / 365) * (exp(y^(1/3)) - 1)
```

@@

\begin{equation*}\label{eq:g}
g(t) = \log^3(1 + 365 t), \quad 0 \le t \le 1,
\end{equation*}

where $t$ is age measured in *years*. 
Its graph is shown in Figure \@ref(fig:figgraph).

```{r figgraph, fig.cap = "The log-cube transform g."}
##g <- function(x) (log(1 + x))^3  # Note: days!
x <- seq(0, 1, length = 1000)
y <- g(x)
plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, col = "blue")
xat <- c(0, 28 / 365, 1 / 2, 1)
yat <- g(xat)
axis(1, at = xat, labels = round(365 * xat, 3))
axis(2, at = yat, labels = round(yat, 1), las = 2)
box()
lines(c(30, 30), c(0, g(30)), lty = 2, col = "darkgreen")
lines(c(0, 30), c(g(30), g(30)), lty = 2, col = "darkgreen")
abline(h = 0, v = 0)
```

The B-P model specifies that, on the transformed age scale, the *exogeneous*
mortality distribution, conditional on death before age one,  is *uniform*, that is, the 
survivor function of $T$ is *linear*.

@@

```{r readdata}
infdat <- readRDS("data/infdat.rds")
```

From the Umeå-Skellefteå region, we follow all persons born in the 
region 1895--1950 until death, reaching the age of 1, or leaving follow-up for whatever 
reason, whichever comes first.

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}
tid <- toTpch(Surv(enter, exit, event) ~ week, data =  infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + x))^3 # x in days.
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

\begin{equation*}\label{eq:g}
g(t) = \log^3(1 + t), \quad 0 \le t \le 365,
\end{equation*}

where $t$ is age measured in *days*. 
Its graph is shown in Figure \@ref(fig:figgraph).

```{r figgraph, fig.cap = "The log-cube transform g."}
##g <- function(x) (log(1 + x))^3  # Note: days!
x <- seq(0, 365, length = 1000)
y <- g(x)
plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, col = "blue")
xat <- c(0, 28, 365)
yat <- g(xat)
axis(1, at = xat)
axis(2, at = yat, labels = round(yat, 1), las = 1)
box()
lines(c(30, 30), c(0, g(30)), lty = 2, col = "darkgreen")
lines(c(0, 30), c(g(30), g(30)), lty = 2, col = "darkgreen")
abline(h = 0, v = 0)
abline(a = 0, b = 1, col = "red")
```

The B-P model specifies that, on the transformed age scale, the *exogeneous*
mortality distribution, conditional on death before age 365 days,  is *uniform*, that is, the 
survivor function of $T$ is *linear*.

@@

```{r readdata}
infdat <- readRDS("data/infdat3.rds")
```

From the Umeå-Skellefteå region, we follow all persons born in the 
region 1895--1950 until death, reaching the age of 1 year, or leaving follow-up for whatever 
reason, whichever comes first. The first few lines (infants) are

```{r infdat3}
source("R/tbl.R")
sinf <- infdat[, c("region", "kon", "birthdate", "deathdate", "exit", "event")]
x <- head(infdat, 5)
tbl(x, caption = "A few infants in the data set.")
```

## The Bourgeois-Pichat procedure

## Add weather data

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}

tid <- toTpch(Surv(enter, exit, event) ~ week, data =  infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```
